In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-24 paper/main.tex
weaken intro sentence

% Main findings and outlook
By exploring this high-dimensional parameter space, we propose that
the balance between CSF production and intracranial pulsatility is key
to shaping the large-scale features of intracranial enrichment
patterns, with the potential to span a wide range of individual and
cohort variability. Moreover, we predict that CSF production, cardiac-
and respiratory pulsatility is not sufficient to explain early
perivascular enrichment, but that fluid flow induced by low-frequency
vasomotion in surface periarterial spaces (on the order of 10 \textmu
m/s) is sufficient, even in the absence of a structural
compartmentalization of the PVS. Conversely, enlarged PVSs in the SAS
will cause a substantial reduction in cardiac- and vasomotion-driven
flow velocities, strongly delay perivascular transport, and thus
impair intracranial enrichment. These findings transfer, reconcile,
integrate and extend insights from clinical, experimental, and
theoretical studies, and lay a new foundation for in-silico studies of
personalized intrathecal drug delivery and brain clearance.

@@

The pulsatile flow of CSF in the SAS and ventricular system,
associated with the cardiac and respiratory cycles, substantially
enhances molecular diffusion through
dispersion~\cite{taylor1953dispersion, watson1983diffusion,
  asgari2016glymphatic, sharp2019dispersion, ray2021quantitative,
  troyetsky2021dispersion}. To account for these dispersive effects,
while bridging from the pulsatile flow time scale of seconds to a
molecular transport time scale of hours, we estimate cardiac- and
respiratory dispersion factors ($R_c$ and $R_r$) in $\Omega_{\rm CSF}$
(\Cref{fig:csf}G, I). To compute these spatially-varying factors, we
combine computational estimates of peak CSF pressure gradients
(\Cref{fig:csf}F, H) with theoretical estimates for shear-augmented
(Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion,
  sharp2019dispersion} (see Supplementary information S1.3 \emph{Estimating
dispersion factors from pulsatile CSF flow} for a complete
description). Combining the cardiac and respiratory effects, we set $R
= R_c + R_r$ in $\Omega_{\rm CSF}$.

@@

We give a concise summary of the numerical approach here (see
Supplementary methods for a comprehensive description). To
solve~\eqref{eq:multi_transport} numerically, we use the interior
penalty discontinuous Galerkin (DG) method with weighted averages and
upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low
regularity properties. Thus, one can only expect lower-order error
convergence near $\Lambda_a$ and $\Lambda_v$, see the references
\cite{masri2024discontinuous} and \cite{laurino2019derivation} for the
numerical analysis of DG and continuous Galerkin schemes,
respectively. However, away from these 1D networks, almost optimal
approximation is
expected~\cite{masri2023discontinuous,koppl2016local}. To numerically
solve the Stokes equations to compute the CSF flow velocity and
pressure in the ventricular system and SAS, we use finite element
spaces that preserve the incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when subsequently
solving~\eqref{eq:multi_transport}\cite{cesmelioglu2022compatible}.
The spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}.
The correctness and accuracy of the numerical solutions were verified by a series of
numerical verification experiments (see Supplementary information S1.5).
The majority of the simulation time was spend on the Stokes flow and transport model simulations. Each Stokes flow simulation utilizes a hybrid parallel approach with 16 MPI processes and 4 threads each, requiring approximately 300 GB of RAM and completing in 23 minutes. The transport model is limited to serial execution by FEniCS\textsubscript{ii}, but employs 16 threads for the linear system factorization, taking 22 minutes and consuming about 50 GB of memory on a AMD EPYC 9684X machine.

>>>>

% Main findings and outlook
By exploring this high-dimensional parameter space, we propose that
the balance between CSF production and intracranial pulsatility is key
to shaping the large-scale features of intracranial enrichment
patterns, with the potential to span a wide range of individual and
cohort variability. Moreover, we predict that CSF production, cardiac-
and respiratory pulsatility is not sufficient to explain early
perivascular enrichment, but that fluid flow induced by low-frequency
vasomotion in surface periarterial spaces (on the order of 10 \textmu
m/s) is sufficient, even in the absence of a structural
compartmentalization of the PVS. Conversely, enlarged PVSs in the SAS
will cause a substantial reduction in cardiac- and vasomotion-driven
flow velocities, strongly delay perivascular transport, and thus
impair intracranial enrichment. These findings help to integrate and extend insights from clinical, experimental, and theoretical studies, and provide a framework for future in-silico investigations of personalized intrathecal drug delivery and brain solute transport.

@@

The pulsatile flow of CSF in the SAS and ventricular system,
associated with the cardiac and respiratory cycles, substantially
enhances molecular diffusion through
dispersion~\cite{taylor1953dispersion, watson1983diffusion,
  asgari2016glymphatic, sharp2019dispersion, ray2021quantitative,
  troyetsky2021dispersion}. To account for these dispersive effects,
while bridging from the pulsatile flow time scale of seconds to a
molecular transport time scale of hours, we estimate cardiac- and
respiratory dispersion factors ($R_c$ and $R_r$) in $\Omega_{\rm CSF}$
(\Cref{fig:csf}G, I). To compute these spatially-varying factors, we rely on a simplified model that combines computational estimates of peak CSF pressure gradients
(\Cref{fig:csf}F, H) with theoretical estimates for shear-augmented
(Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion,
  sharp2019dispersion} (see Supplementary information S1.3 \emph{Estimating
dispersion factors from pulsatile CSF flow} for a complete
description). Combining the cardiac and respiratory effects, we set $R
= R_c + R_r$ in $\Omega_{\rm CSF}$.

@@

We give a concise summary of the numerical approach here (see
Supplementary methods for a comprehensive description). To
solve~\eqref{eq:multi_transport} numerically, we use the interior
penalty discontinuous Galerkin (DG) method with weighted averages and
upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low
regularity properties. Thus, one can only expect lower-order error
convergence near $\Lambda_a$ and $\Lambda_v$, see the references
\cite{masri2024discontinuous} and \cite{laurino2019derivation} for the
numerical analysis of DG and continuous Galerkin schemes,
respectively. However, away from these 1D networks, almost optimal
approximation is
expected~\cite{masri2023discontinuous,koppl2016local}. To numerically
solve the Stokes equations to compute the CSF flow velocity and
pressure in the ventricular system and SAS, we use finite element
spaces that preserve the incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when subsequently
solving~\eqref{eq:multi_transport}\cite{cesmelioglu2022compatible}.
The spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}.
The correctness and accuracy of the numerical solutions were verified
by a series of numerical verification experiments (see Supplementary
information S1.5).

The majority of the simulation time was spent on the Stokes flow and transport model simulations. Each Stokes flow simulation utilizes a hybrid parallel approach with 16 MPI processes and 4 threads each, requiring approximately 300 GB of RAM and completing in 23 minutes. The transport model is limited to serial execution by FEniCS\textsubscript{ii}, but employs 16 threads for the linear system factorization, taking 22 minutes and consuming about 50 GB of memory on a AMD EPYC 9684X machine.
